---
title: "Exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database."
---


##Synopsis: 

The weather data used for this report is available with NOAA started spanning from year 1950 and end in November 2011. It is said that in the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


Historically, the TORNADO is by far the most deadly and hurtful weather event. However, in average, the most deadly is the heat (on the form of HEAT and EXCESSIVE HEAT event types) and the most wounding is ICE STORM followed by FLOOD and heat (HEAT and EXCESSIVE HEAT).

For crop damages, the three most damaging events have been the EXCESSIVE WETNESS, COLD AND WET CONDITIONS and DAMAGING FREEZE, whereas for the property damages they are HEAVY RAIN/SEVERE WEATHER and TORNADOES, TSTM WIND, HAIL. However, the reported frequency of occurrence of these events is very low (lesser than 5), which is not enough for charactering them as probables, without disdain of its devastating nature.

For events with a frequency greater that seven, the DROUGHT event causes in average the most severe damages for the crops, nearly followed, perhaps paradoxically, by the RIVER FLOOD. Regarding the property damages, the hurricane is by far the most costly event.

In the last 17 years (1984-2011) the damage occasioned by weather events tends to grow, what reinforces, but does not conclusively show, the suspicions about global climate changes


Environment Information
---

```{r myEnv, echo=TRUE}
        sessionInfo()
```
              
Loading Required Libraries 
---

``` {r loadlibraries, echo=TRUE}
        library(knitr)
        library(downloader)
        library(lubridate)
        library(plyr)
        library(ggplot2)
        library(scales)
        library(reshape2)
```



Setting Rawdata directory
---

```{r additionaldirs, echo=TRUE}
        if (!file.exists("rawData"))  {dir.create("rawData") }
```


Data Processing
---
###Loading and Shaping Data###
The data is in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. 
The download file is available at [Storm Data approx(47MB)](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)


```{r Loaddata, echo=TRUE, cache=FALSE}
        if (!file.exists("repdata-data-StormData.csv.bz2"))  
        { 
                 fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2" 
                 download(fileUrl,"repdata-data-StormData.csv.bz2" ,mode="wb")
        }

        stormdf <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), header=TRUE, sep=",")

## Shaping Labels
        names(stormdf) <- toupper(names(stormdf))
        names(stormdf) <- gsub("__","CODE",names(stormdf)) 
        names(stormdf) <- gsub("LONGITUDE_","LONGITUDE_E",names(stormdf))
        names(stormdf) <- gsub("BGN_LOCATI","BGN_LOCATION",names(stormdf))


        stormSubsetdf <- subset(stormdf, FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG > 0, select = c(1:8, 22:28,32:33))
        stormSubsetdf$BGN_DATE <- mdy_hms(stormSubsetdf$BGN_DATE)
```


Data Processing
---


####Processing of population health data####
Reshaping and melting population health data for plotting
```{r healthdata, echo=TRUE, cache=FALSE}

        total_FI <- ddply(stormSubsetdf, .(EVTYPE), summarise, 
                        total_FATALITIES=sum(FATALITIES, na.rm = TRUE), 
                        total_INJURIES = sum(INJURIES, na.rm = TRUE), 
                        mean_FATALITIES = mean(FATALITIES, na.rm = TRUE), 
                        mean_INJURIES = mean(INJURIES, na.rm = TRUE), 
                        sd_FATALITIES = sd(FATALITIES, na.rm = TRUE), 
                        sd_INJURIES =  sd(INJURIES, na.rm = TRUE))
#
        top10FI_total <- head(total_FI[order(-total_FI$total_FATALITIES,total_FI$total_INJURIES) , ] , 10)
        top10_fata <- head(total_FI[order(total_FI$total_FATALITIES, decreasing = TRUE), c("EVTYPE", "total_FATALITIES")] , 10)
        top10_inju <- head(total_FI[order(total_FI$total_INJURIES, decreasing = TRUE), c("EVTYPE", "total_INJURIES")] , 10)

        top10_FInew <- merge(top10_fata, top10_inju, by="EVTYPE", all=TRUE)
        melted_FI <- melt(top10_FInew, id.vars=c("EVTYPE"), measure.vars=c("total_FATALITIES","total_INJURIES"), preserve.na = FALSE)

# colnames(top10state_fata) [3] <- "SUM"
# colnames(top10state_inju) [3] <- "SUM"
# top10_FIrbind <- rbind(top10state_fata,top10state_inju)

```





####Processing of economic data####
Methodology:


The data fields used to dertermine economic values are _PROPDMG, PROPDMGEXP, CROPDMG and CROPDMGEXP_.
We assume that PROPDMGEXP and CROPDMGEXP corresponds to exponent. The economic values can be determined by multiplying PROPDMG by PROPDMGEXP and CROPDMG by CROPDMGEXP.

Resolving exponential values:  

The PROPDMGEXP has levels  `r levels(stormSubsetdf$PROPDMGEXP)`   ,whereas CROPDMGEXP has   `r levels(stormSubsetdf$CROPDMGEXP)` 
we have to drive the cost by using  The methodology i used to 
According to [NATIONAL WEATHER SERVICE: Estimated Damage](http://www.nws.noaa.gov/wsom/manual/archives/NF429405.HTML#2.2.5  Estimated) _Alphabetical characters used to signify magnitude include K for thousands, M for millions, B for billions, and T for trillions_. In addtionl for the values like "1" "2" "3" "4" "5" "6" "7" "8" will be used as exponential value. Others such as ""  "-" "?" "+" "0" will be treated as 1 so that the actual value shall not be tempered.

```{r numaricValue, echo=TRUE, cache=FALSE}

stormSubsetdf$PROPCOST <-  stormSubsetdf$PROPDMG *      ifelse(toupper(stormSubsetdf$PROPDMGEXP) == "H", 100    , 
                                                        ifelse(toupper(stormSubsetdf$PROPDMGEXP) == "K", 1000   , 
                                                        ifelse(toupper(stormSubsetdf$PROPDMGEXP) == "M", 1e+06  , 
                                                        ifelse(toupper(stormSubsetdf$PROPDMGEXP) == "B", 1e+09  ,  
                                                        ifelse((stormSubsetdf$PROPDMGEXP == "?" | 
                                                                stormSubsetdf$PROPDMGEXP == "+" | 
                                                                stormSubsetdf$PROPDMGEXP == "+" | 
                                                                stormSubsetdf$PROPDMGEXP == "0"| 
                                                                stormSubsetdf$PROPDMGEXP == ""), 1              ,
                                                        10**!is.na(stormSubsetdf$PROPDMGEXP))))))  

stormSubsetdf$CROPCOST <-  stormSubsetdf$CROPDMG *      ifelse(toupper(stormSubsetdf$CROPDMGEXP) == "H", 100    , 
                                                        ifelse(toupper(stormSubsetdf$CROPDMGEXP) == "K", 1000   , 
                                                        ifelse(toupper(stormSubsetdf$CROPDMGEXP) == "M", 1e+06  , 
                                                        ifelse(toupper(stormSubsetdf$CROPDMGEXP) == "B", 1e+09  ,  
                                                        ifelse((stormSubsetdf$CROPDMGEXP == "?" | 
                                                                stormSubsetdf$CROPDMGEXP == "+" | 
                                                                stormSubsetdf$CROPDMGEXP == "+" | 
                                                                stormSubsetdf$CROPDMGEXP == "0"| 
                                                                stormSubsetdf$CROPDMGEXP == ""), 1              ,
                                                        10**!is.na(stormSubsetdf$CROPDMGEXP))))))  
```


Reshaping and melting economic consequences data for plotting
```{r economicdata, echo=TRUE, cache=FALSE}


        total_PC <- ddply(stormSubsetdf, .(EVTYPE), summarise, 
                        total_PROPCOST  = sum(PROPCOST, na.rm = TRUE), 
                        total_CROPCOST  = sum(CROPCOST, na.rm = TRUE), 
                        mean_PROPCOST   = mean(PROPCOST, na.rm = TRUE), 
                        mean_CROPCOST   = mean(CROPCOST, na.rm = TRUE), 
                        sd_PROPCOST     = sd(PROPCOST, na.rm = TRUE), 
                        sd_CROPCOST     = sd(CROPCOST, na.rm = TRUE))
#
        top10PC_total <- head(total_PC[order(-total_PC$total_PROPCOST,total_PC$total_CROPCOST) , ] , 10)
        top10_prop <- head(total_PC[order(total_PC$total_PROPCOST, decreasing = TRUE), c("EVTYPE", "total_PROPCOST")] , 10)
        top10_crop <- head(total_PC[order(total_PC$total_CROPCOST, decreasing = TRUE), c("EVTYPE", "total_CROPCOST")] , 10)


        top10_PCnew <- merge(top10_prop, top10_crop, by="EVTYPE", all=TRUE)
        melted_PC <- melt(top10_PCnew, id.vars=c("EVTYPE"), measure.vars=c("total_PROPCOST","total_CROPCOST"), preserve.na = FALSE)

```


####Processing of population health data for the State####

```{r healthdata_state, echo=TRUE, cache=FALSE}

        totalstate_FI <- ddply(stormSubsetdf, .(EVTYPE,STATE), summarise, 
                        FATALITIES=sum(FATALITIES, na.rm = TRUE), 
                        INJURIES = sum(INJURIES, na.rm = TRUE))
#
        top10stateFI_total <- head(totalstate_FI[order(-totalstate_FI$FATALITIES,totalstate_FI$INJURIES) , ] , 10)
        top10state_fata <- head(totalstate_FI[order(totalstate_FI$FATALITIES, decreasing = TRUE), c("EVTYPE", "STATE", "FATALITIES")] , 10)
        top10state_inju <- head(totalstate_FI[order(totalstate_FI$INJURIES, decreasing = TRUE), c("EVTYPE", "STATE", "INJURIES")] , 10)


# top10state_FInew <- merge(top10state_fata, top10State_inju, by="EVTYPE", all=TRUE)
# Statefata_sum <- aggregate(list(sumFATA=totalstate_FI$total_FATALITIES), list(STATE=totalstate_FI$STATE), sum, rm.na=TRUE)
```

```{r economicdata_state, echo=TRUE, cache=FALSE}

        totalstate_PC <- ddply(stormSubsetdf, .(EVTYPE,STATE), summarise, 
                        PROPERTY  = sum(PROPCOST, na.rm = TRUE), 
                        CROP  = sum(CROPCOST, na.rm = TRUE))
#
        top10statePC_total <- head(totalstate_PC[order(-totalstate_PC$PROPERTY,totalstate_PC$CROP) , ] , 10)
        top10state_prop <- head(totalstate_PC[order(totalstate_PC$PROPERTY, decreasing = TRUE), c("EVTYPE", "STATE", "PROPERTY")] , 10)
        top10state_crop <- head(totalstate_PC[order(totalstate_PC$CROP, decreasing = TRUE), c("EVTYPE", "STATE", "CROP")] , 10)


# top10state_PCnew <- merge(top10state_prop, top10State_crop, by="EVTYPE", all=TRUE)
# Stateprop_sum <- aggregate(list(sumPROP=totalstate_PC$total_PROPCOST), list(STATE=totalstate_PC$STATE), sum, rm.na=TRUE)

```


Results
---


####Top 10 most harmful events across the United States with respect to polulation health####

The table below provides some useful information for more detail work 

```{r FI_table, echo=TRUE}
#grid.table(top10FI_total, show.rownames=FALSE)
        top10FI_total[, c(1:3)]
```


```{r top10health, echo=TRUE, cache=FALSE}

        ggplot(melted_FI, aes(EVTYPE, value, fill=variable)) + geom_bar(stat="identity", position="dodge") + theme(axis.text.x = element_text(angle= 90, hjust = 1)) + labs(x="Total Fatalities and Injuries", y="Event Type", title="Top 10 Weather events across United States\neffecting population health") + scale_y_continuous(labels = comma) + scale_fill_discrete(name="Population\nHealth", labels=c("Fatalities", "Injuries"))
```


####Top 10 most harmful events across the United States with respect to economic consequences####

The table below provides some useful information for more detail work 

```{r PC_table, echo=TRUE}
# grid.table(top10PC_total, show.rownames=FALSE)
        top10PC_total[, c(1:3)]
```


```{r top10economic, echo=TRUE, cache=FALSE}
        ggplot(melted_PC, aes(EVTYPE, value, fill=variable)) + geom_bar(stat="identity", position="dodge") + theme(axis.text.x = element_text(angle= 90, hjust = 1)) + labs(x="Property and Crop Damage", y="Event Type", title="Top 10 Weather events across United States\ncausing economic consequences") + scale_y_continuous(labels = dollar) + scale_fill_discrete(name="Economic\nConsequences", labels=c("Property Damage", "Crop Damage"))
```


```{r,echo=TRUE}

top10stateFI_total
```
As you know Illinois has a widely varying climate. Most of Illinois has a humid continental climate with hot, humid summers and cool to cold winters.The all-time high temperature was 117 F (47 C), recorded on July 14, 1954, at East St. Louis, while the all time low temperature was -36 F (-38 C), recorded on January 5, 1999, at Congerville. A temperature of -37 F (-39 C), was recorded on January 15, 2009, at Rochelle.

```{r,echo=TRUE}

top10statePC_total
```
All types of floods can occur in California, though 90% are caused by riverine flooding.[1] Such flooding generally occurs as a result of excessive rainfall, excessive snowmelt, excessive runoff, levee failure or a combination of these sources.
Big Floods such as:
January 1982: Northern California flood
1986 California and Western Nevada floods

```{r statesY, echo=TRUE}
        totalstate_FI_Y <- ddply(stormSubsetdf, .(YEAR = format(BGN_DATE, "%Y"), EVTYPE,STATE), summarise, 
                               FATALITIES=sum(FATALITIES, na.rm = TRUE), 
                               INJURIES = sum(INJURIES, na.rm = TRUE))
        
        totalstate_PC_Y <- ddply(stormSubsetdf, .(YEAR = format(BGN_DATE, "%Y"), EVTYPE,STATE), summarise, 
                                PROPERTY  = sum(PROPCOST, na.rm = TRUE), 
                                CROP  = sum(CROPCOST, na.rm = TRUE))

        top10state_fataY  <- head(totalstate_FI_Y[order(-totalstate_FI_Y$FATALITIES),c(1,2,3,4) ] ,10)
        top10state_injuY  <- head(totalstate_FI_Y[order(-totalstate_FI_Y$INJURIES),c(1,2,3,5) ] ,10)      
        top10state_propY  <- head(totalstate_PC_Y[order(-totalstate_PC_Y$PROPERTY),c(1,2,3,4) ] ,10)
        top10state_cropY  <- head(totalstate_PC_Y[order(-totalstate_PC_Y$CROP),c(1,2,3,5) ] ,10)
        
        total_Y <- ddply(stormSubsetdf, .(YEAR = format(BGN_DATE, "%Y")), summarise, 
                         FATALITIES=sum(FATALITIES, na.rm = TRUE), 
                         INJURIES = sum(INJURIES, na.rm = TRUE))

        top10_FI_Y <- head(total_Y[order(-total_Y$FATALITIES, -total_Y$INJURIES, decreasing=FALSE ),  ] ,10)
        top10_PC_Y <- head(total_Y[order(-total_Y$PROPERTY, -total_Y$CROP, decreasing=FALSE ),  ] ,10)
```